Constructing a universe for the setoid model

### 2020-05-11 paper.tex
Add content to paper.tex

The first step is therefore to give up the idea of defining equality as a single
inductive definition, and adopt the approach according to which equality should
be defined in a specific way for each type former in the type theory. That is,
every type former should come equipped with its own notion of equality, from
which a general notion of equality is derived.

Two systems of ITT with extensional properties that follow this approach are OTT
and cubical TT. In the first one ... TODO. In cubical TT ... TODO. Note that we
do not consider HoTT since, although it is a version of ITT with extensional
properties, it's not known to enjoy canonicity.

The system that we will consider in this paper follows the same approach of OTT,
namely the setoid model... TODO

* Internal encoding of setoids

* Setoid hell

@@

@@

@@

@@

@@

Following this, we seen that we can actually encode the universe above as an
inductive type $\ad{in-U}$ that \emph{carves out} all types in $\Set$ that are
in the image of $\El$:
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \Set\} \{B : A \to \Set\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}
Using this $\U$ and $\El$ can be given as follows:
\begin{align*}
  \U & = \Sigma (A : \Set)\ (\textsf{in-U}\ A) \\
  \El & = \textsf{proj}_1
\end{align*}

@@

To provide such a universe we need to define a setoid $\setoidU : \Setoid$ and a
family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
formers like $\Pi$-types and booleans. This can be obtained as an inductive-recursive type.
\todo{expand definition to include more details}
%
\begin{align*}
  & \textsf{data} \ \setoidU : \Set_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \El : \setoidU \to \Set \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}

As said before, however, we would like to avoid having to assume
inductive-recursive types, of more generally induction-recursion, in the
metatheory.

Our result is that a modified form of this translation also works for
the more complex inductive-recursive type we need to model the universe of
setoids. 
%
In particular, in addition to $\ad{in-U}$ for defining $\U$ as before, we also
introduce a family $\ad{in-U∼}$ of binary relations between types in the
universe, from which we then define $\_\sim_{\setoidU}\_$.
%
However, we now need an inductive-inductive type in the target theory:
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1 \\
  & \setoidU : \Set_1 \\
  & \El : \setoidU \to \Set \\
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \_\vdash\_\sim_{\El}\_ : \{a\ a' : \U\} \to a\ \sim_{\setoidU}\ a' \to \El\ a \to \El\ a' \to \Prop \\
  & \\
  & \setoidU = \Sigma\ (X : \Set)\ (\ad{in-U}\ X) \\
  & \El = \textsf{proj}_1 \\
  & (X , p) \sim_{\setoidU} (X' , p') =
   \ \parallel \Sigma\ (R : X \to X' \to \Prop)\ (\ad{in-U∼}\ p\ p'\ R) \parallel
\end{align*}

where $\parallel A \parallel\ : \Prop$ denotes the propositional truncation of $A :
\Set$. Intuition would suggest to define $\_\vdash\_\sim_{\El}\_$ by projecting
the relation out of the proof of $a \sim_{\setoidU} a'$, in much the same way as
$\El$ is defined by projecting out of $\U$. That is, $p \vdash x \sim_{\El} y =
\ad{proj}_1\ p\ x\ y$.
%
However, this doesn't work since the type of $p$ is propositionally truncated,
hence it cannot be used to construct a proof-relevant object. We can work around
this by instead defining $\_\vdash\_\sim_{\El}\_$ by induction on the
codes $a\ a' : \setoidU$, explicitly for each type former, mutually with a proof
that this definition is logically equivalent to what we would obtain if we could
project out of the truncation.

>>>>

The first step is therefore to give up the idea of defining equality as a single
inductive definition, and adopt the approach according to which equality should
be defined specifically for each type former in the type theory. That is, every
type former $A$ should be introduced together with its own notion of equality
$A^\sim$, and the internal notion of identity in type theory should refer to
that relation.
%
If one does this, then funext can be obtained by simply defining equality on
function types to be point-wise equality. That is, $(A \to B)^\sim\ f\ g :\equiv
\forall (x : A) , B^\sim\ (f x)\ (g x)$.

* One possible answer: setoids

This form of extensional reasoning is already available in type theory via
\emph{setoids}. A setoid is just a type $A$ (in this setting, we assume that all
types are sets in the HoTT sense) together with a propositional equivalence
relation $A^\sim : A \to A \to \Prop$.
%
One can thus reason extensionally in ITT by defining an internal notion of
setoid, and working with those instead of normal sets/types. Note that this
approach only works provided one ensures that every construction performed on
setoids respects the setoid relations involved. Unfortunately doing so is not an
easy task, and most constructions beyond the more trivial ones will generate a
great amount of proof obligations, especially when dependent setoids come into
play, leading to a situation commonly known as \emph{setoid hell}.

@@

The good news is that all this work proving that our type theoretic
constructions respect the setoid relations is not necessary, because these
already respect setoid equality \emph{by construction}. More precisely, there
exists a sound interpretation of MLTT (with pi, sigma, etc...) into setoids,
that maps closed types/contexts to setoids and dependent types to indexed
setoids.
%
Thus there is no need to explicitly work with setoids and setoid relations in
TT, as one can simply carry out their constructions in plain MLTT and use the
aforementioned interpretation to map that construction into the world of setoid.

This is interesting in principle, but practically useless. Since our
constructions still take place in MLTT, we can't take advantage of the features
of the setoid model our syntactic constructions are mapped to, since they are
not accessible internall, from within the system itself.
%
This is where a key property of models comes into play. Models inform consistent
extensions of the type theory, based on what holds true in the model, by
reflecting the semanics into the syntax via additional syntactic constructs that
are then mapped back to the original notions via the semantics.

In our case, the existence of a setoid model suggests us ways to extend the
syntax of MLTT with syntactic constructs that correspond to the setoid
relations, allowing us to access and use them when proving theorems within the
theory.
%
This is the idea behind Setoid Type Theory, SeTT for short.

* Two incarnations of ITT that follow this approach are OTT and cubical TT. In OTT
... TODO. In cubical TT ... TODO. Note that we do not include HoTT in the
discussion since, although it is a version of ITT with extensional properties,
it's not yet known to enjoy canonicity.

@@

%% The system that we will consider in this paper follows the same motivation of
%% OTT, of giving a syntactic presentation of the setoid model in [cite].

@@

\todo{Say a bit about how one can turn the setoid model to a setoid translation
  (strict model and such...)}

\todo{This gives a translation/explanation/implementation of extensionality in
  terms of intensional TT}

@@

\todo{Mention that the universe cannot be univalent}

\todo{Maybe mention the fact that it should be possible to give an elimination
  principle for the setoid universe. This also makes sense since it's not univalent.}

@@

Following this, we seen that we can actually encode the universe above as an
inductive type $\ad{in-U}$ that \emph{carves out} all types in $\Set$ that are
in the image of $\El$. In other words, $\ad{in-U}$ is a predicate that holds for
any type that would have been obtained via $\El$ in the usual
inductive-recursive definition. As $\El$ is indexed on the type of codes, the
definition of $\ad{in-U}$ quite expectedly reflecs the inductive structure of
codes.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \Set\} \{B : A \to \Set\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}
Using this $\U$ and $\El$ can be given as follows:
\begin{align*}
  \U & = \Sigma (A : \Set)\ (\textsf{in-U}\ A) \\
  \El & = \textsf{proj}_1
\end{align*}

@@

To provide such a universe we need to define a setoid $\setoidU : \Setoid$ and a
family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
formers like $\Pi$-types and booleans. This can be defined as the following
inductive-recursive type. \todo{expand definition to include more details}
%
\begin{align*}
  & \textsf{data} \ \setoidU : \Set_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \El : \setoidU \to \Set \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}

The point of having $\setoidU$ and $\_\sim_\setoidU\_$ is quite clear: our
setoid universe is a type, hence itself a setoid, hence described as any other
setoid by a carrier set and an equivalence relation on it. These functions are
achieved by $\setoidU$ and $\_\sim_{\setoidU}\_$, respectively.

For any code $A$ in the universe, we want a type $\El\ A$; this, in turn, should
be a setoid as well. In other words, we want a family of setoids indexed over
setoids. We thus need to explain, for any $A$, what is the carrier and
equivalence relation of $\El\ A$. This is achieved by $\El : \setoidU \to \Set$,
and $\_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to
\El\ a \to \El\ a' \to \Prop$. Note that we need this to be indexed over code
equality, because El is a type family indexed over U, and hence it must respect
the equality on U. \todo{rewrite in better form}

As already pointed out, however, we would like to avoid having to assume
inductive-recursive types, of more generally induction-recursion, in the
metatheory.

Our result is that a modified form of this translation also works for
the more complex inductive-recursive type we need to model the universe of
setoids. 
%
In particular, in addition to $\ad{in-U}$ for defining $\U$ as before, we also
introduce a family $\ad{in-U∼}$ of binary relations between types in the
universe, from which we then define $\_\sim_{\setoidU}\_$.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1
\end{align*}

Note that $\ad{in-U}$ appears in the sort of $\ad{in-U∼}$, thus making this
definition as instance of induction-induction [cite]. \todo{add brief
  explanation of induction-induction}.

As in the previous example, we can now define $\setoidU$ as a $\Sigma$ type, and
$\El$ as the corresponding first projection.

\begin{align*}
  & \setoidU : \Set_1 \\
  & \setoidU = \Sigma\ (X : \Set)\ (\ad{in-U}\ X) \\
  & \El : \setoidU \to \Set \\
  & \El = \textsf{proj}_1
\end{align*}

What is left now is to define the setoid equality relation on the universe, as
well as the setoid equality relation on $\El\ A$ for any $A$ in $\setoidU$.
%
The definition of $\_\sim_{\setoidU}\_$ follows, in principle, that of
$\setoidU$. A ...

Similarly, two codes $A, B$ in the universe $\setoidU$ are equal when there
exists a setoid equivalence relation on their respective sets $\El\ A$ and $\El
B$. Intuitively, since elements of a setoid are only ever compared to elements
of the same setoid, this is only possible if $A, B$ are codes for the same
setoid, that is, if $A \sim_{\setoidU} B$. Existence and well-formedness of such
relations is expressed via the type $\ad{in-U~}$ just defined, hence formally we
should have something like:

\[
A \sim_{\setoidU} B :\equiv \Sigma (R: ...) ...
\]

or, using pattern matching:

\[
(A, p) \sim_{\setoidU} (A', p') :\equiv \Sigma (R: ...) ...
\]

Unfortunately this definition only captures the idea we are aiming for, but does
not actually typecheck. In fact, $\_\sim_{\setoidU}\_$ is a $\Prop_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition, but $\Sigma (R: ...)
...$ clearly isn't.

The solution is actually quite simple, and it just involves truncating the
$\Sigma$ type described above to force it to be in $\Prop_1$. More precisely, we
assume to have in our metatheory, in addition to a hierarchy of universes of
strict propositions $\Prop_i$, a truncation operation $\parallel\_\parallel :
\Set_i \to \Prop_i$ with constructor $|\_| : \{A : \Set_i\} \to A \to \parallel
A \parallel$ and eliminator $\ad{withTrunc} : \{B : \Prop_i\} \to (A \to B) \to
\parallel A \parallel \to B$. Note that truncation can be defined in Agda
extended with the universe of strict propositions, with the eliminator replaced
by pattern matching.

\begin{align*}
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & (X , p) \sim_{\setoidU} (X' , p') =
   \ \parallel \Sigma\ (R : X \to X' \to \Prop)\ (\ad{in-U∼}\ p\ p'\ R) \parallel
\end{align*}

This trick might seem to simple to the point of being naive, but as we will see
it is sufficient to make the entire definition go through.

We are now left to define $\_\vdash\_\sim_{\El}\_$.

\begin{align*}
  & \_\vdash\_\sim_{\El}\_ : \{a\ a' : \U\} \to a\ \sim_{\setoidU}\ a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}

Intuition would suggest to define $\_\vdash\_\sim_{\El}\_$ by projecting
the relation out of the proof of $A \sim_{\setoidU} A'$, in much the same way as
$\El$ is defined by projecting out of $\U$. That is, $p \vdash x \sim_{\El} y =
\ad{proj}_1\ p\ x\ y$.
%
However, this doesn't work since the type of $p$ is propositionally truncated,
hence it cannot be used to construct a proof-relevant object. In other words,
the proof $p$ cannot be eliminated into $El A \to El A' \to Prop$, because
$El A \to El A' \to Prop$ is not a proposition, but a $\Set$.

We can work around this limitation by defining $p \vdash x \sim_{\El} y$ by
induction on the codes $a\ a' : \setoidU$, in a way that is logically equivalent
to the proposition we would have obtained by $\ad{proj}_1\ p\ x\ y$ if that
expression was well-typed.

Note that for this definition to go through, we sometimes need the fact that the
relation we are building is indeed equivalent to the one we would have obtained
if we could project out of the truncation. That is, we need an additional proof
that for any concrete $R$ and $\ad{inR}$, $| (R, \ad{inR}) | \vdash x \sim_{\El}
y \iff R\ x\ y$. We thus need to define $\_\vdash\_\sim_{\El}\_$ mutually with a proof
of the following

\[
...
\]
